The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-09-29 paper.tex
Revisao "Related Work"

In the context of embedded systems, the size of \code{trail\_t} is typically 
only 3 bytes (1 byte for each field), imposing a negligible memory overhead 
even for trails that only await a single event and terminate.
%
For instance, the \emph{CTP} collection protocol ported to \CEU reaches eight 
simultaneous lines of execution with an overhead of 2\% in comparison to the 
original version in \emph{nesC} (a dialect of \emph{C} for event-driven 
programming)~\cite{ceu.sensys13}.

@@

    event void Boot.booted () {
        ceu_go_init();
#ifdef CEU_WCLOCKS
        call Timer.startPeriodic(10);
#endif
    }

#ifdef CEU_WCLOCKS
    event void Timer.fired () {
        ceu_go_wclock(10000);
    }
#endif

#ifdef _EVT_PHOTO_READDONE
    event void Photo.readDone (uint16_t val) {
        ceu_go_event(EVT_PHOTO_READDONE, (void*)val);
    }
#endif

#ifdef _EVT_RADIO_SENDDONE
    event void RadioSend.sendDone (message_t* msg) {
        ceu_go_event(EVT_RADIO_SENDDONE, msg);
    }
#endif

#ifdef _EVT_RADIO_RECEIVE
    event message_t* RadioReceive.receive (message_t* msg) {
        ceu_go_event(EVT_RADIO_RECEIVE, msg);
        return msg;
    }
#endif

@@

TODO: refs/cite WSNs

As an example, Figure~\ref{lst.impl.tinyos} shows our binding for \emph{TinyOS} 
which maps \emph{nesC} callbacks to input events in \CEU.
%
The file \code{ceu.h} (included in line 3) contains all definitions for the 
compiled \CEU program, which are further queried through \code{\#ifdef}'s.
The file \code{ceu.c} (included in line 4) contains the main loop of \CEU 
pointing to the labels defined in the program.
The callback \code{Boot.booted} (lines 6--11) is called by TinyOS on startup, 
so we initialize \CEU inside it (line 7).
If the \CEU program uses timers, we also start a periodic timer (lines 8--10) 
that triggers callback \code{Timer.fired} (lines 13--17) every 10 milliseconds 
and advances the wall-clock time of \CEU (line 15)%
\footnote{We also offer a mechanism to start the underlying timer on demand to
avoid the ``battery unfriendly'' 10ms polling.}.
The remaining lines map pre-defined TinyOS events that can be used in \CEU 
programs, such as the light sensor (lines 19--23) and the radio transceiver 
(lines 25--36).

@@

\CEU has a strong influence from Esterel~\cite{esterel.ieee91} and embraces the 
disciplined synchronous-reactive model with support for lexical composition of 
lines of execution.
%
However, there are fundamental semantic differences that prevents the design of 
\CEU as pure extensions to Esterel (i.e., on top of its formal semantics).
%
In particular, Esterel has a notion of time similar to that of digital circuits 
in which multiple signals can be active at a clock tick.
In fact, Esterel is also used in hardware design.
%
In \CEU, instead of clock ticks, the occurrence of a single external event that 
defines a time unit.
The event-driven approach is popular in many 
communities~\cite{TOS,games,GUI,callbacks,nodejs}, and is essentially 
synchronous, even though it is not commonly/formally 
noticed/acknowledged/recognized as such.
(TODO: rewrite)
%
We believe that for software design, this approach is more familiar to 
programmers and simplifies the reasoning about concurrency.
For instance, the uniqueness of external events in \CEU is a prerequisite for 
the temporal analysis that enables safe shared-memory concurrency.
%
Finally, \CEU also distinguishes external events from stack-based internal 
events, which provides a limited form of coroutines that supports \CEU's 
reactive statements.

>>>>

In the context of embedded systems, the size of \code{trail\_t} is typically 
only 3 bytes (1 byte for each field), imposing a negligible memory overhead 
even for trails that only await a single event and terminate.
%
For instance, the \emph{CTP} collection protocol ported to \CEU reaches eight 
simultaneous lines of execution with an overhead of 2\% in comparison to the 
original version in \emph{nesC}~\cite{wsn.nesc} (a dialect of \emph{C} for 
event-driven programming)~\cite{ceu.sensys13}.

@@

    event void Boot.booted () {
        ceu_sys_init();
#ifdef CEU_WCLOCKS
        call Timer.startPeriodic(10);
#endif
    }

#ifdef CEU_WCLOCKS
    event void Timer.fired () {
        int dt = 10000;
        ceu_sys_go(EVT__WCLOCK, &dt);
    }
#endif

#ifdef _EVT_PHOTO_READDONE
    event void Photo.readDone (uint16_t val) {
        ceu_sys_go(EVT_PHOTO_READDONE, &val);
    }
#endif

#ifdef _EVT_RADIO_SENDDONE
    event void RadioSend.sendDone (message_t* msg) {
        ceu_sys_go(EVT_RADIO_SENDDONE, &msg);
    }
#endif

#ifdef _EVT_RADIO_RECEIVE
    event message_t* RadioReceive.receive (message_t* msg) {
        ceu_sys_go(EVT_RADIO_RECEIVE, &msg);
        return msg;
    }
#endif

@@

As an example, Figure~\ref{lst.impl.tinyos} shows our binding for 
\emph{TinyOS}~\cite{wsn.tos}, which maps callbacks to input events in \CEU.
%
The file \code{ceu.h} (included in line 3) contains all definitions for the 
compiled \CEU program, which are further queried through \code{\#ifdef}'s.
The file \code{ceu.c} (included in line 4) contains the main loop of \CEU 
pointing to the labels defined in the program.
The callback \code{Boot.booted} (lines 6--11) is called by TinyOS on startup, 
so we initialize \CEU inside it (line 7).
If the \CEU program uses timers, we also start a periodic timer (lines 8--10) 
that triggers callback \code{Timer.fired} (lines 13--18) every 10 milliseconds 
and advances the wall-clock time of \CEU (line 16)%
\footnote{We also offer a mechanism to start the underlying timer on demand to
avoid the ``battery unfriendly'' 10ms polling.}.
The remaining lines map pre-defined TinyOS events that can be used in \CEU 
programs, such as the light sensor (lines 20--24) and the radio transceiver 
(lines 26--37).

@@

\CEU has a strong influence from Esterel~\cite{esterel.ieee91} and embraces the 
disciplined synchronous-reactive model with support for lexical composition of 
lines of execution.
%
However, there are fundamental semantic differences that prevents the design of 
\CEU as pure extensions to Esterel (i.e., on top of its formal semantics).
%
In particular, Esterel has a notion of time similar to that of digital circuits 
in which multiple signals can be active at a clock tick.
In fact, Esterel is also used in hardware design.
%
In \CEU, instead of clock ticks, the occurrence of a single external event that 
defines a time unit.
%
\CEU also distinguishes external events from stack-based internal events, which 
provides a limited form of coroutines that supports reactive statements (e.g., 
\code{await} and \code{par/or}).

The event-driven approach~\cite{sync_async.whynotthreads} is well known and 
popular in many software communities,
such as web frameworks in the client side (e.g., \emph{jQuery}~\cite{js.jquery} 
and \emph{Rx.js}~\cite{js.rx}) and server side (e.g., 
\emph{Node.js}~\cite{js.node} and \emph{SEDA}~\cite{sync_async.seda}),
GUI toolkits (e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java 
Swing}~\cite{java.swing}),
and Games~\cite{gamepatterns}.
%
Like \CEU, event-driven programming is essentially synchronous, i.e., events go 
through a queue and are dispatched sequentially and atomically to prevent race 
conditions.
%
We believe that for software design, this approach is more familiar to 
programmers and simplifies the reasoning about concurrency.
For instance, the uniqueness of external events in \CEU is a prerequisite for 
the temporal analysis that enables safe shared-memory concurrency.
